Deep Neural Networks for Bilingual Lexicon Extraction

### 2014-06-07 paper/main.tex
Clean up paper LaTeX, begin extending for long paper version

%\author{Jon Gauthier\affilA\,and\,Arthur Tsang\affilB \\
%  \affilA Symbolic Systems, \affilB Computer Science \\
%Stanford University, Stanford, CA 94035 \\
%{\tt \{jgauthie,atsang2\}@stanford.edu}}

@@

\begin{abstract}
  We explore a method for producing bilingual lexicons from non-parallel
  corpora. In contrast with most previous approaches, our method separates the steps
  of \begin{inparaenum}[(1)]
    \item constructing distributed word representations from monolingual
      corpora, and
    \item building a mapping between these representations.
  \end{inparaenum}
  This gives us precise control over the independent development of the word representations and the mapping between them.
  %By performing these two steps separately, we gain great flexibility in determining how our word representations are constructed. % are able to construct distributed word representations with greater flexibility. % TODO more specific and contentful
  We demonstrate that a deep neural network can effectively learn a translational
  mapping between these representations for the purposes of bilingual lexicon extraction more accurately than a linear model.
  %explain baseline
  %We suggest that a deep learning model beats a linear baseline by allowing more flexibility. % make this more concrete
\end{abstract}

@@

Bilingual lexicon extraction is obviously of most utility when applied to
under-resourced language pairs which do not already benefit from a surplus of
lexicon materials. These same language pairs, lacking resources as simple as quality translation dictionaries, also lack sufficient amounts of manually aligned
text to train BLE models. % briefly mention that each individual language may have a significant corpus though?
For this reason, most recent work on this task focuses on
minimally supervised approaches which do not require parallel corpora
\cite{rapp1995,peirsman2010}.

@@

The present paper is motivated by the hypothesis that this standard bilingual
vector space approach to BLE is inherently limited by the design of the space
itself.
% SHORT We suggest that the construction of the space along axes composed of a hand-picked set of bilingual seeds has the potential to limit the expressiveness of the word representations contained therein.
Consider, for example, a vector
space constructed in this fashion using just two seeds: \textit{hot} and
\textit{cold}. Such a space may be effective at representing the difference
between words like \textit{fire} and \textit{ice}. However, with these two axes
it would fail to capture many other distinctions, such as the difference between
the words \textit{up} and \textit{down}. The quality of the vector space is
thus heavily dependent on the quality of its seeds.

\newcite{mikolov2013exploiting} suggest an alternate method for bilingual lexicon
extraction, centered around two monolingual vector spaces $E$ (corresponding to the
source language) and $F$ (corresponding to the target language). These vector spaces
are constructed using a standard neural language model \cite{mikolov2013}. We see this as a promising alternative to the aforementioned bilingual vector space approach, as it allows for the construction of robust distributed representations whose quality is not dependent on manually selected seeds. The authors find
that a linear model trained via stochastic gradient descent to map between these two
vector spaces can effectively map source-language word representations to the
target-language word vectors corresponding to their translations.

We extend this work in attempting to model a \emph{translation function} $t : E \to F$.
% SHORT
% We capitalize on the fact that these vector spaces may be independently constructed, experimenting with different configurations for each space.
In this paper we approximate the transformation $t$ between the two spaces using a deep neural
network, trained to accept as input word vectors from the source language space
$E$ and output word vectors in the target language space $F$.

@@

Our translation algorithm makes use of two monolingual corpora: one consisting of text
in the source language and the other containing text in the target language. We presume
the corpora have the following properties:
\subparagraph{Large and broad-domain.} With this property given, we can safely assume
that for any common
word which appears in the source-language corpus, we will also find its
translation somewhere in the target-language corpus.
If large corpora are unavailable, we can make use of smaller narrow-domain corpora, such as religious texts.
\subparagraph{Non-parallel.} This means that our algorithm does not depend on any sort
of alignment, lexical or topical. This relative flexibility allows us to demonstrate
that the method can be applied in situations where parallel corpora are not available.

@@

In addition to the two monolingual corpora, we train on a minimal
%, manually-created
set of word translation pairs.
% TODO I don't like "such as a swadesh list" --- Jon
The specifics of these training pairs are described in \Cref{subsec:translation-nn}.

%In addition to using the corpora, we also rely on a small seed set of several hundred English % -Spanish word pairs to train the translation function.
% Include the seed sets as an appendix?

@@

\Cref{fig:translation} illustrates the process of translating a single word from a source
language to a target language (here English and Spanish, respectively) in this model. We
begin with an input source-language word and retrieve a corresponding pre-computed word
representation in the source VSM. We use this source-language word vector as an input to
a neural network, which outputs a vector in the target-language VSM. We perform a simple
nearest-neighbor search in the target-language VSM to find the closest existing word
vector, and take its corresponding target-language word to be the translation of the
original source-language word.

@@

For each of the two corpora introduced in \Cref{sec:data}, we train a continuous vector space model on 10-word contexts, using an implementation in the \textit{Gensim} software package of \texttt{word2vec}, a neural language model \cite{rehurek2010,mikolov2013}. This neural language model attempts to learn word representations which minimize error in the task of predicting a word's context given the word itself. The resulting word vectors are thus theoretically optimal (i.e., minimum-cost) representations of the meanings underlying their corresponding words. This is in contrast with the word representations built by a standard BLE approach, which define themselves in terms of manually constructed axes, and thus do not necessarily account for a maximal amount of variance in meaning among the words represented.

We experiment with different word vector dimensionalities, and detail our specific choices later in \Cref{subsec:test-data}.

%A VSM is structured such that the relative position of a word encodes aspects of its meaning.
%For example, similar words map into similar neighborhoods, and addition by the same vector can produce analogies (citation needed).\unnec
% Ok, I suppose we don't need to give that much detail (especially for a 4 page paper). A brief mention, though, can help explain the intuition that some structure can be extrapolated by a neural network --- Arthur
% Since independently-produced VSMs maintain similar patterns, this suggests we can build a translation function from a small number of seeds.

@@

The computation central to the process of translation is the transformation of a
source-language word representation into a target-language word representation via
a deep neural network. We learn this transformation from a small set of training pairs.
%broad-domain training pairs, each mapping a single source language word to a single
%translationally equivalent
%target language word.

% TODO added this.. what do you think? --- Jon
The standard bilingual vector space approach detailed in the introduction would
base its word representations on these training pairs. In contrast with this method,
our model uses these translation pairs only in its final learning step, while training
the neural network which models the translation function. For this reason, the model
can benefit from effective word representations developed independently from the
translation system itself. Furthermore, it can provide direct feedback through a training cost function
on the quality of the translation pairs.

We investigate several different methods 
of producing translation pairs.
In line with previous
approaches, we fill a set
of translation pairs
\texttt{Cog} with cognates shared between the two
languages \cite{koehn2002}. We also build a small separate set
\texttt{ConcNN}
consisting of concrete nouns, and attempt to restrict the set to word pairs for which
both the source- and target-language words have just one sense in their respective
languages.

@@

We compute the linear transformation by taking $A = W_F W_E^+$, where $W_E^+$ is the Moore--Penrose
pseudoinverse of $W_E$.
% SHORT
% It is necessary to use the pseudoinverse of $W_E$, as the matrix is unlikely to be square and invertible.
Then given any source-language word vector $w_e$, we calculate
the transformed target-language word vector as $w_f = A w_e$.

\section{Evaluation}
\label{sec:evaluation}
% TODO maybe for long paper: note that test sets mirror the style of the seed set. e.g.
% model trained on concrete nouns is tested on unseen concrete nouns; model tested on
% cognates is tested on unseen cognates
We evaluate our translation model on an unseen set of English-Spanish word pairs. Since our model builds a ranked list of nearest-neighbor translation candidates, we use the mean reciprocal rank (MRR) of the correct translation within the list:
\begin{equation}
    \textit{MRR} = \frac{1}{n}\sum_{j=1}^n \frac{1}{r_j}
\end{equation}
where $r_j$ is the rank of the correct translation for pair $j$ out of $n$.
%For example, if every correct translation is ranked second highest, $\textit{MRR} = \frac{1}{2}$. % leaving out for space considerations

@@

We establish three separate test environments, each defined by a choice of training data,
input corpus format and VSM design. The vector space dimensionalities given below
were derived through trial and error. % SHORT We find that different VSM dimensionalities are optimal for different corpus inputs. Overall, we see that performance varies significantly based on these test environment properties.
The environments are as follows:
\begin{description}
\item[\texttt{Cog}:] Minimally preprocessed corpora, \texttt{Cog} training data, 500-dimensional word vectors % TODO long paper: 5000/750 min word threshold
% Perhaps note that we don't do any Levenshtein mumbo-jumbo? --- Arthur
\item[\texttt{ConcNN}:] Minimally preprocessed corpora, \texttt{ConcNN} training data, 500-dimensional word vectors % TODO long paper: 10000/1500 min word threshold
\item[\texttt{CogLem}:] Lemmatized Spanish corpus, \texttt{Cog} training data, 1000-dimensional word vectors % TODO long paper: 2500/375 min word threshold
\end{description}

@@

Our evaluation metric looks for a single possible translation of a test input word in
ranked output. As we explain in \Cref{subsec:error-analysis}, this may needlessly penalize
the algorithm.
% SHORT , as it often suggests reasonable translations earlier in the ranked output which happen to not match our strict search requirements.
For example, our algorithm is
tested on a cognate pair $(\textit{concept}, \textit{concepto})$, and is penalized for
suggesting the translation \textit{idea}, a suitable synonym of the desired target word.
A fourth test run \bt{CogFix} considers the algorithm's output on the \texttt{Cog} dataset,
and accepts such sensible alternate translations. % SHORT (rather than stipulating in this case that all source-language cognates translate to their corresponding target-language cognates).

@@

The best absolute performance is achieved by a neural network with 250 hidden-layer
neurons in the \bt{CogFix} training environment. Interestingly, the hidden layer of this model is of
lower dimensionality than the test input and output word vectors (each 500 dimensions).
% TODO repetitive "This suggests" --- Jon
% Mention that the Linear model may be overfitting, if compressing first leads to better results? --- Arthur
This suggests that the model learns some useful compressed intermediate representation of
the input source-language word in the translation training process.
% SHORTCAND
Its MRR shows us that on average the algorithm suggests an acceptable translation in the second or third position of the ranked list of possible translations.

@@

While the translation algorithm fails to predict the exactly correct translation a
significant amount of the time, an analysis of common prediction errors is encouraging.
We identify several common error patterns in the
algorithm's output in \Cref{tbl:errors}, and elaborate on the classes of errors below:
\subparagraph{Synonyms.} With many of the "errors" of this form, the algorithm provided an acceptable alternate translation that we simply did
not expect. These cases are treated as successes in the \bt{CogFix} training environment, as described
in \Cref{subsec:results}.
\subparagraph{Near misses.} In these cases the algorithm mapped into a precise neighborhood
of related words, but picked the wrong element from the set. % SHORT
% For example, the output word \textit{chocolate} is strongly related to the expected word \textit{vainilla}.
% TODO long paper: future work might examine the actual distance values and "give up" if the top few words are all very close
\subparagraph{Right neighborhood, wrong spot.} We occasionally mapped into a correct larger
area of the target language vector space, but still made a significant semantic error.
% SHORT
% With the input word \textit{election}, for example, the algorithm predicts as translations target-language words related to political campaigns.
\subparagraph{Alternate senses.} In several cases on the \texttt{Cog} training
set and its derivatives, the algorithm learned to predict translations of the input words which
corresponded to valid alternate interpretations of the source language word.
% SHORT
% For example, it interprets the English \textit{division} in the sense of a military unit (rather than in the mathematical sense as we expected).
Note that these cases were not treated as
successes in the \bt{CogFix} training environment. We hope to eliminate this class of errors
in future work by using multi-prototype VSMs (see \Cref{sec:conclusion}).

@@

For simplicity, our current VSM design associates a single word vector with each word
in the source or target language corpus. However, training error could be greatly
reduced by allowing polysemous words to be associated with multiple word vectors
\cite{reisinger2010,huang2012}. We believe such an extension will remove much of the remaining
ambiguity in the input vector spaces.

>>>>

\author{Jon Gauthier,\affilA{} Arthur Tsang,\affilB{} Christopher Potts\affilC{} \\
  \affilA{} Symbolic Systems Program, \affilB{}Department of Computer Science, \affilC{}Department of Linguistics \\
Stanford University, Stanford, CA 94035 \\
{\tt \{jgauthie,atsang2,cgpotts\}@stanford.edu}}

@@

\begin{abstract}
  We explore a method for producing bilingual lexicons from non-parallel
  corpora. In contrast with most previous approaches, our method separates the
  steps of \begin{inparaenum}[(1)]
  \item constructing distributed word representations from monolingual corpora,
    and
  \item building a translational mapping.
  \end{inparaenum}

  This gives us precise control over the independent development of the word
  representations and the mapping between them.

  We demonstrate that a deep neural network can effectively learn a
  translational mapping between these representations for the purposes of
  bilingual lexicon extraction more accurately than a closely related linear
  model.
\end{abstract}

@@

Bilingual lexicon extraction is of most utility when applied to under-resourced
language pairs which do not already benefit from a surplus of lexicon materials.
These same language pairs, lacking resources as simple as quality translation
dictionaries, also lack sufficient amounts of manually aligned text to train BLE
models. For this reason, most recent work on this task focuses on minimally
supervised approaches which do not require parallel corpora
\cite{rapp1995,peirsman2010}.

@@

The present paper is motivated by the hypothesis that this standard bilingual
vector space approach to BLE is inherently limited by the design of the space
itself. We suggest that the construction of the space along axes composed of a
hand-picked set of bilingual seeds has the potential to limit the expressiveness
of the word representations contained therein. Consider, for example, a vector
space constructed in this fashion using just two seeds: \textit{hot} and
\textit{cold}. Such a space may be effective at representing the difference
between words like \textit{fire} and \textit{ice}. However, with these two axes
it would fail to capture many other distinctions, such as the difference between
the words \textit{up} and \textit{down}. The quality of the vector space is thus
heavily dependent on the quality of its seeds.

\newcite{mikolov2013exploiting} suggest an alternate method for bilingual
lexicon extraction, centered around two monolingual vector spaces $E$
(corresponding to the source language) and $F$ (corresponding to the target
language). These vector spaces are constructed using a standard neural language
model~\cite{mikolov2013}. This is a promising alternative to the aforementioned
bilingual vector space approach, as it allows for the construction of robust
distributed representations whose quality is not dependent on manually selected
seeds. The authors find that a linear model trained via stochastic gradient
descent to map between these two vector spaces can effectively find word-level
translations from the source to the target language. We extend this work in
attempting to approximate a \emph{translation function} $t : E \to F$ using a
deep neural network, trained to accept as input word vectors from the source
language space $E$ and output word vectors in the target language space $F$. We
capitalize on the fact that these vector spaces may be independently
constructed, experimenting with different configurations for each space.

@@

Our translation algorithm makes use of two monolingual corpora: one consisting
of text in the source language and the other containing text in the target
language. We require only that the corpora have the following properties:

\subparagraph{Large and broad-domain.} With this property given, we can safely
assume that for any common word which appears in the source-language corpus, we
will also find its translation somewhere in the target-language corpus. If large
corpora are unavailable, we can make use of smaller narrow-domain corpora (e.g.,
religious texts, famous novels).

\subparagraph{Non-parallel.} This means that our algorithm does not depend on
any sort of alignment, lexical or topical. This relative flexibility allows us
to demonstrate that the method can be applied in situations where parallel
corpora are not available.

@@

In addition to the two monolingual corpora, we train on a minimal set of word
translation pairs. The specifics of these training pairs are described in
\Cref{subsec:translation-nn}.

@@

\Cref{fig:translation} illustrates the process of translating a single word from
a source language to a target language (here English and Spanish, respectively)
in this model. We begin with an input source-language word and retrieve a
corresponding pre-computed word representation in the source VSM\@. We use this
source-language word vector as an input to a neural network, which outputs a
vector in the target-language VSM\@. We perform a simple nearest-neighbor search
in the target-language VSM to find the closest existing word vector, and take
its corresponding target-language word to be the translation of the original
source-language word.

@@

For each of the two corpora introduced in \Cref{sec:data}, we train a continuous
vector space model on 10-word contexts, using an implementation in the
\textit{Gensim} software package of \texttt{word2vec}, a neural language
model~\cite{rehurek2010,mikolov2013}. This neural language model attempts to
learn word representations which minimize error in the task of predicting a
word's context given the word itself. The resulting word vectors are thus
theoretically optimal (i.e., minimum-cost) representations of the meanings
underlying their corresponding words. This is in contrast with the word
representations built by a standard BLE approach, which define themselves in
terms of manually constructed axes, and thus do not necessarily account for a
maximal amount of variance in meaning among the words represented. We experiment
with different word vector dimensionalities, and detail our specific choices
later in \Cref{subsec:test-data}.

While we focus on constructing a neural language model with \texttt{word2vec},
we believe any distributional model of word meaning would suffice for this
task~\cite{bengio2003,turian2010,collobert2011}.

@@

The central computation is the transformation of a source-language word
representation into a target-language word representation via a deep neural
network. We learn this transformation from a small set of training pairs.

The standard bilingual vector space approach detailed in the introduction would
base its word representations on these training pairs. In contrast with this
method, our model uses these translation pairs only in its final learning step,
while training the neural network which models the translation function. For
this reason, the model can benefit from effective word representations developed
independently from the translation system itself. Furthermore, it can provide
direct feedback through a training cost function on the quality of the
translation pairs.

We investigate several different methods of producing translation pairs. In line
with previous approaches, we fill a set of translation pairs \texttt{Cog} with
cognates shared between the two languages~\cite{koehn2002}. We also build a
small separate set \texttt{ConcNN} consisting of concrete nouns, and attempt to
restrict the set to word pairs for which both the source- and target-language
words have just one sense in their respective languages.

@@

We compute the linear transformation by taking $A = W_F W_E^+$, where $W_E^+$ is
the Moore-Penrose pseudoinverse of $W_E$. It is necessary to use the
pseudoinverse of $W_E$, as the matrix is unlikely to be square and invertible.
Then given any source-language word vector $w_e$, we calculate the transformed
target-language word vector as $w_f = A w_e$.

\section{Evaluation}
\label{sec:evaluation}
We evaluate our translation model on an unseen set of English-Spanish word
pairs. These test sets are customized to match the algorithm's training
environment. For example, a translation algorithm trained on concrete nouns from
the training set \texttt{ConcNN} will be tested on other unseen concrete nouns.
Since our model builds a ranked list of nearest-neighbor translation candidates,
we can calculate the mean reciprocal rank (MRR) of the correct translation
within the list:
\begin{equation}
    \textit{MRR} = \frac{1}{n}\sum_{j=1}^n \frac{1}{r_j}
\end{equation}
where $r_j$ is the rank of the correct translation for pair $j$ out of $n$. For
example, a mean reciprocal rank of 1 would indicate that every translation
output ranked the correct translation in the first position. A mean reciprocal
rank of $\frac{1}{2}$ would indicate that the average translation output placed
the correct translation as the second most likely in a ranked list.

@@

We establish three separate test environments, each defined by a choice of
training data, input corpus format, and VSM design. The vector space
dimensionalities given below were derived through trial and error. We find that
different VSM dimensionalities are optimal for different corpus inputs. Overall,
we see that performance varies significantly based on these test environment
properties. The environments are as follows:
\begin{description}
\item[\texttt{Cog}:] Minimally preprocessed corpora, \texttt{Cog} training data,
  500-dimensional word vectors % TODO long paper: 5000/750 min word threshold
\item[\texttt{ConcNN}:] Minimally preprocessed corpora, \texttt{ConcNN} training
  data, 500-dimensional word
  vectors % TODO long paper: 10000/1500 min word threshold
\item[\texttt{CogLem}:] Lemmatized Spanish corpus, \texttt{Cog} training data,
  1000-dimensional word vectors % TODO long paper: 2500/375 min word threshold
\end{description}

@@

Our evaluation metric looks for a single possible translation of a test input
word in ranked output. As we explain in \Cref{subsec:error-analysis}, this may
needlessly penalize the algorithm, as it often suggests reasonable translations
earlier in the ranked output which happen to not match our strict search
requirements. For example, our algorithm is tested on a cognate pair
$(\textit{concept}, \textit{concepto})$, and is penalized for suggesting the
translation \textit{idea}, a suitable synonym of the desired target word. A
fourth test run \bt{CogFix} considers the algorithm's output on the \texttt{Cog}
dataset, and accepts such sensible alternate translations (rather than
stipulating in this case that all source-language cognates translate to their
corresponding target-language cognates).

@@

The best absolute performance is achieved by a neural network with 250
hidden-layer neurons in the \bt{CogFix} training environment. Interestingly, the
hidden layer of this model is of lower dimensionality than the test input and
output word vectors (each 500 dimensions). This suggests that the model learns
some useful compressed intermediate representation of the input source-language
word in the translation training process. Its MRR shows us that on average the
algorithm suggests an acceptable translation in the second or third position of
the ranked list of possible translations.

@@

While the translation algorithm fails to predict the exactly correct translation a
significant amount of the time, an analysis of common prediction errors is encouraging.
We identify several common error patterns in the
algorithm's output in \Cref{tbl:errors}, and elaborate on the classes of errors below:

\subparagraph{Synonyms.} With many of the "errors" of this form, the algorithm
provided an acceptable alternate translation that we simply did not expect.
These cases are treated as successes in the \bt{CogFix} training environment, as
described in \Cref{subsec:results}.

\subparagraph{Near misses.} In these cases the algorithm mapped into a precise
neighborhood of related words, but picked the wrong element from the set. For
example, the output word \textit{chocolate} is strongly related to the expected
word \textit{vainilla}. % TODO long paper: future work might examine
% the actual distance values and "give up" if the top few words are all very
% close

\subparagraph{Right neighborhood, wrong spot.} We occasionally mapped into a
correct larger area of the target language vector space, but still made a
significant semantic error. With the input word \textit{election}, for example,
the algorithm predicts as translations target-language words related to
political campaigns.

\subparagraph{Alternate senses.} In several cases on the \texttt{Cog} training
set and its derivatives, the algorithm learned to predict translations of the
input words which corresponded to valid alternate interpretations of the source
language word. For example, it interprets the English \textit{division} in the
sense of a military unit (rather than in the mathematical sense as we expected).
These cases were not treated as successes in the \bt{CogFix} training
environment. We hope to eliminate this class of errors in future work by using
multi-prototype VSMs (see \Cref{sec:conclusion}).

@@

For simplicity, our current VSM design associates a single word vector with each
word in the source or target language corpus. However, training error could be
greatly reduced by allowing polysemous words to be associated with multiple word
vectors~\cite{reisinger2010,huang2012}. We believe such an extension will remove
much of the remaining ambiguity in the input vector spaces.
